Making Sequential Programs Deterministic

### 2019-04-05 java-design.tex

Our implementation treats arrays analogously to collections:
the semantics and rules are similar even though the syntax varies.
For simplicity this
paper uses the term ``collection'' and the type \<Collection> to represent
collections and arrays.
Collections include any class (including user-defined classes) that
implements the \<Iterable> or \<Iterator> interfaces; this includes all
Java collections such as \<List>s and \<Set>s.

@@

We noticed the following properties with respect to determinism regarding the most widely used \<Map> implementations:
\begin{itemize}
    \item \<HashMap> can never guarantee fully deterministic iteration over its entries. A \codeid{Det HashMap} does not exist.
    \item \<LinkedHashMap> can have any of \<Det>, \<OrderNonDet>, or \<NonDet> types. Iterating over a \<LinkedHashMap> returns
    its entries in the order of their insertion. An \<OrderNonDet LinkedHashMap> can be created by calling its constructor with a \<HashMap>.
    \item \<TreeMap> can either be \<Det> or \<NonDet>. An \<OrderNonDet TreeMap> doesn't exist because the entries are always ordered.
\end{itemize}

We prohibit the creation of a \codeid{@Det HashMap} or an \codeid{@OrderNonDet TreeMap} in our determinism checker.
\Cref{fig-creation-rules} formalizes the type wellformedness rules for maps and sets. Similar constraints apply to \<HashSet>s and \<TreeSet>s which are
also expressed by the last two rules of \cref{fig-creation-rules}.

\begin{figure}
    $\infer[\rulename{valid hashmap}]{\vdash: \|\kappa\ HashMap()|}{\kappa \in \{ \<OrderNonDet>, \<NonDet> \}}$

@@

    $\infer[\rulename{valid treemap}]{\vdash: \|\kappa\ TreeMap()|}{\kappa \in \{ \<Det>, \<NonDet> \}}$

@@

    $\infer[\rulename{valid hashset}]{\vdash: \|\kappa\ HashSet()|}{\kappa \in \{ \<OrderNonDet>, \<NonDet> \}}$

@@

    $\infer[\rulename{valid treeset}]{\vdash: \|\kappa\ TreeSet()|}{\kappa \in \{ \<Det>, \<NonDet> \}}$
    \caption{\<Map> and \<Set> creation rules}
    \todo{Are these type well-formedness rules?}
    \todo{What about \<new TreeSet(myOrderedCollection<Det ...>)> which yields a
        deterministic set?}
    \label{fig-creation-rules}
\end{figure}

In the JDK, \<LinkedHashSet> is implemented as a subclass of \<HashMap>.
\todo{Talk about this design choice. Josh Bloch's book: "If the code surrounding the first declaration depended on LinkedHashSet's ordering policy,
    then it would be incorrect to substitute HashSet for LinkedHashSet in the declaration, because HashSet makes no guarantee concerning iteration order."}

>>>>

Our implementation treats arrays analogously to collections:
the semantics and rules are similar even though the syntax varies.
For simplicity this
paper uses the term ``collection'' and the type \<Collection> to represent
collections and arrays.
Collections include any type (including user-defined classes) that
implements the \<Iterable> or \<Iterator> interfaces; this includes all
Java collections such as \<List>s and \<Set>s.
Our implementation also supports any type that implements the \<Map>
interface.

@@

The most widely used \<Map> implementations have the following properties:
\begin{itemize}
    \item \<HashMap> is implemented in terms of a hash table, which never
      guarantees deterministic iteration over its entries. A \codeid{@Det
        HashMap} does not exist.
    \item \<LinkedHashMap>, like \<List>, can have any of the \<@NonDet>,
      \<@OrderNonDet>, or \<@Det> type qualifiers. Iterating over a
      \<LinkedHashMap> returns
    its entries in the order of their insertion. An \<OrderNonDet
    LinkedHashMap> can be created by calling its constructor with a
    \<@OrderNonDet HashMap>.
    \item \<TreeMap> can either be \<@Det> or \<@NonDet>. An \<@OrderNonDet
      TreeMap> doesn't exist because the entries are always ordered.
\end{itemize}

\TheDeterminismChecker prohibits the creation of a \codeid{@Det HashMap} or an \codeid{@OrderNonDet TreeMap}.
\Cref{fig-creation-rules} formalizes the type well-formedness rules for
Java \<Map>s and \<Set>s.

\begin{figure}
    $\infer[\rulename{valid hashmap}]{\vdash: \|\kappa\ \<HashMap>|}{\kappa \in \{ \<OrderNonDet>, \<NonDet> \}}$

@@

    $\infer[\rulename{valid treemap}]{\vdash: \|\kappa\ \<TreeMap>|}{\kappa \in \{ \<Det>, \<NonDet> \}}$

@@

    $\infer[\rulename{valid hashset}]{\vdash: \|\kappa\ \<HashSet>|}{\kappa \in \{ \<OrderNonDet>, \<NonDet> \}}$

@@

    $\infer[\rulename{valid treeset}]{\vdash: \|\kappa\ \<TreeSet>|}{\kappa \in \{ \<Det>, \<NonDet> \}}$
    \caption{Type well-formedness rules for subclasses of \<Map> and \<Set>.}
    \label{fig-creation-rules}
\end{figure}

The JDK implements \<LinkedHashSet> as a subclass of \<HashMap>\todo{should
  this be HashSet?  It's not true as written.}
in order to inherit parts of \<HashMap>'s implementation.
\todo{Talk about this design choice. Josh Bloch's book: "If the code surrounding the first declaration depended on LinkedHashSet's ordering policy,
    then it would be incorrect to substitute HashSet for LinkedHashSet in the declaration, because HashSet makes no guarantee concerning iteration order."} 
\todo{The above is not clear.  What are the consequences for our
  implementation.  What goes wrong or is hard to annotate or is surprising?
  I'm not sure what is the point of this paragraph.}
